Make any stupid plot you want

### 2021-05-19 draft/sections/math.tex
monoid section

\subsubsection{Measurement Scales: Monoid Actions}
\label{sec:math:data:monoid}
Implementing expressive visual encodings requires formally describing the structure on the components of the fiber. We conjecture that the structure can be formally defined as the actions of a monoid on the component. We can also discuss complex structures since a core property of monoids is composability \cite{yorgeyMonoidsThemeVariations}. In doing so, we can specify the properties of the component that must be preserved in a graphic representation.

A monoid \cite{Monoid2021} $\monoid$ is a set with a binary operation $\ast:\monoid \times \monoid\rightarrow \monoid$ that satisfies the axioms:
\begin{align*}
    \textbf{associativity}\; & \text{for all} a, b, c \in \monoid \, (a \bullet b) \bullet c = a \bullet (b \bullet c)\\
    \textbf{identity}\; & \text{for all } a\in \monoid,\,  e\bullet a = a 
\end{align*} 
As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,nlabauthorsAction2021} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
\begin{align*}
    \textbf{associativity}\;& \text{for all } f,g \in \monoid_i \text{ and } x\in \dfiber_i,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
    \textbf{identity}\;& \text{for all } x\in \dfiber_i, e\in \monoid_i,\,  e\bullet x = x 
\end{align*}
We conjecture that given these properties, if there is a partial ordering on both sides of the action then the action is equivariant and monotonic.

\begin{figure}[H]
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=\textwidth]{figures/math/monoid_total.png}
        \caption{}
        \label{fig:math:data:monoid:total}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{figures/math/monoid_partial.png}
        \caption{}
        \label{fig:math:data:monoid:partial}
    \end{subfigure}
    \caption{The action $\bullet$ in \autoref{fig:math:data:monoid:total} is the arrows from the partial order diagram of weather states on the left to the diagram of weather states on the right. The action maps the weather states to themselves, thereby preserving the ordering defined by the monoid $\ast$ on both sides of the action. We conjecture that the action in \autoref{fig:math:data:monoid:partial} is monotonic when the partial ordering of the weather states is the same as the partial order of the elements they are mapped to. Given $sun \geq clouds\geq rain,\, lightining$ on the right, we conjecture that the action  $sun,\, clouds \rightarrow clouds$, and $rain,\, lightining \rightarrow rain$ is structure preserving such that the relative ordering of elements is the same as the elements they are mapped to.}
    \label{fig:math:data:monoid}
\end{figure}

One example of monoids are partial orderings on a set, such as seen in \autoref{fig:math:data:monoid}. Each hasse diagram of the set of weather states describes an ordering on the set; the arrow goes from the lesser value to the greater one. For example, $storm \leq rain$. In \autoref{fig:math:data:monoid}, the action $\bullet$ maps the elements of a set of weather states into itself by mapping them into other elements of the weather states. The action in \autoref{fig:math:data:monoid:total}, represented as the arrows between the hasse diagrams of the weather states, maps the weather states to themselves; therefore the ordering of the weather states is identical on both sides of the action. The action $\bullet$ in \autoref{fig:math:data:monoid:partial} is a monotone map \cite{fongInvitationAppliedCategory2019}
\begin{equation*}
    if\, a \leq b \; then \; \bullet(a) \leq \bullet(b) \mid a,b \in \dfiber_{i}
\end{equation*}
where the structure the action preserves is the relative, rather than exact, ordering. Since groups are monoids with invertible operations, this definition of structure is broad enough to include the Steven's measurement scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}.

As with the fiber \dfiber\, the total monoid space \monoid\ is the cartesian product
\begin{equation}
    \label{eq:math:data:monoid:decompose}
\monoid= \monoid_{0} \times \ldots \times \monoid_{i}\times \ldots \times\ldots \monoid_{n}
\end{equation}
of each monoid \(\monoid_{i}\) on \(\dfiber_{i}\).  The monoid is added to the specification of the fiber \((\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)\)

>>>>

\subsubsection{Measurement Scales: Monoid Actions}
\label{sec:math:data:monoid}
Expressiveness of visual encodings is defined as encoding the relations of the data in the visual space \cite{mackinlayAutomatingDesignGraphical1986} and we formally describe these relations as monoid actions on the data component. We describe relations using monoids because they encompass the Steven's measurement scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}, partial order relations, such as multi-ranked indicators \cite{bruggemannRankingPrioritizationMultiindicator2011}, and are composable \cite{yorgeyMonoidsThemeVariations}.

A monoid \cite{Monoid2021} $\monoid$ is a set with a binary operation $\ast:\monoid \times \monoid\rightarrow \monoid$ that satisfies the axioms:
\begin{align}
    \textbf{associativity}\; & \text{for all } m_j, m_k, m_l \in \monoid \, (m_j \ast m_k) \ast m_l = m_j \ast (m_k \ast m_l)\\
    \textbf{identity}\; & \text{for all } m_j\in \monoid,\,  e\ast m_j = m_j 
\end{align} 
As defined on data components $\dfiber$, a left monoid action \cite{SemigroupAction2021,nlabauthorsAction2021} is a set $\dfiber$ with an action $\bullet: \monoid\times \dfiber \rightarrow \dfiber$ with the properties:
\begin{align}
    \textbf{associativity}\;& \text{for all } m_j, m_k \in \monoid \text{ and } x\in \dfiber,\, m_j\bullet(m_k\bullet x) = (m_j\ast m_k) \bullet x\label{eq:math:data:monoid:asso}\\
    \textbf{identity}\;& \text{for all } x\in \dfiber,\; e\in \monoid,\,  e\bullet x = x \label{eq:math:data:monoid:identity}
\end{align}

As with the fiber \dfiber\, the total monoid space \monoid\ is the cartesian product
\begin{equation}
    \label{eq:math:data:monoid:decompose}
\monoid= \monoid_{0} \times \ldots \times \monoid_{i}\times \ldots \times\ldots \monoid_{n}
\end{equation}
of each monoid \(\monoid_{i}\) on each component \(\dfiber_{i}\).  The monoid is added to the specification of the fiber \((\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)\)

As defined in \autoref{eq:math:data:monoid:asso}, the functions $m_j, m_k \in \monoid_i$ are composable
\begin{equation}
\begin{tikzcd}
    \dfiber_i \arrow[d, "m_j"'] \arrow[rd, "m_j\;\circ\; m_k"] &         \\
    \dfiber_i \arrow[r, "m_k"']  & \dfiber_i
\end{tikzcd}
\end{equation}
This means either applying $m_k$ to the elements in $F_i$ and then $m_j$ to the results or composing $m_j\circ m_k$ and applying the composition to the elements in $F_i$ will yield the same result.

\begin{figure}[H]
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=\textwidth]{figures/math/monoid_total.png}
        \caption{}
        \label{fig:math:data:monoid:total}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{figures/math/monoid_partial.png}
        \caption{}
        \label{fig:math:data:monoid:partial}
    \end{subfigure}
    \caption{In this example, the partially ordered set of weather values are the elements of the fiber component $F_i$. In \autoref{fig:math:data:monoid:total} the function $m_j$ is the identity function $e \in \monoid$ that takes every element $x \in \dfiber_i$ to itself. In \autoref{fig:math:data:monoid:partial}, the function $m_k$ maps the weather elements on the left to the elements on the right such that multiple weather elements on the right are mapped into the same element on the left. The function $m_k$ is order preserving, meaning for example that $sun \geq clouds$ and $m_k(sun) \geq m_k(clouds)$ are both true. The functions $m_j, m_k$ are composable, meaning that whether they are applied in stages, such as the output of $e$ \autoref{fig:math:data:monoid:total} is the input to \autoref{fig:math:data:monoid:partial}, or the functions (the arrows) are first composed $m_j \circ m_k$, the result will be the same.}
    \label{fig:math:data:monoid}
\end{figure}
In \autoref{fig:math:data:monoid}, the arbitrarily chosen functions $m_j, m_k\in \monoid_i$ act on the partially ordered set $\dfiber_i=weather$. In this example, we define the functions $m_j, m_k \in \monoid_i$ applied to elements of  $\dfiber_i$ to be monotone maps \cite{fongInvitationAppliedCategory2019}
\begin{equation*}
    \label{eq:math:data:monoid:monotone}
    if\, a \leq b \; then \; m_j(a) \leq m_j(b) \mid a,b \in \dfiber_{i}
\end{equation*}
In \autoref{fig:math:data:monoid:total}, the function $m_j$ is the identity function $e$ which takes every element in $\dfiber_i$ to itself. The function $m_k$ in \autoref{fig:math:data:monoid:partial} maps the elements of weather on the left to the subset on the right in a way that satisfies the monotonicity condition in \autoref{eq:math:data:monoid:monotone}. For example, on the right $sun \geq clouds$ is true and on the left $m_k(sun) \geq m_k(clouds)$ is true since $m_k(sun)=clouds$ and $m_k(clouds)=rains$ and $clouds\geq rain$. Since the functions $m_j, m_k$ are composable, the result is the same whether the input to $m_k$ is the output of $m_j$ or the functions (arrows) are combined before being applied to the elements in $F_i$.
